A Partitioning Algorithm for Maximum Common Subgraph Problems

### 2017-02-17 paper/paper.tex

We initially assume that graphs are unlabelled, undirected and without loops
(\cref{sec:extensions} describes how these restrictions may be relaxed).
The vertices and edge sets of a graph $\graphG$ are denoted $\V(\graphG)$ and $\E(\graphG)$.  The
set of vertices adjacent to vertex $v$ in graph $\graphG$ is called the
\emph{neighbourhood} of $v$, denoted $\N(\graphG, v)$. We denote by $\invN(\graphG, v)$ the
\emph{inverse neighbourhood} of $v$, being the set of the vertices not adjacent
to $v$ (excluding $v$ itself). A \emph{subgraph} of a graph $\graphG$ is a graph
consisting of some of the vertices of $\graphG$, and all of the edges between these
vertices. A \emph{common subgraph} of two graphs is a graph which is
(isomorphic to) a subgraph of two graphs simultaneously, and a \emph{maximum}
common subgraph is one with as many vertices as possible. (In this paper, all
subgraphs are induced. The \emph{maximum common partial subgraph} problem
instead asks for a common non-induced subgraph with as many \emph{edges} as
possible \citep{DBLP:conf/cp/NdiayeS11}.)

@@

\begin{algorithm}[t]
\DontPrintSemicolon
\nl $\FuncSty{search}(\AlgVar{future},M)$ \;
\nl \Begin{
%\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
\nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$\AlgVar{incumbent} \gets M$} \label{StoreIncumbent}
%\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
\medskip
\nl $\AlgVar{bound} \gets |M|  + \sum_{\langle \setG,\setH \rangle \in \AlgVar{future}} \min(|\setG|,|\setH|)$ \label{CalcBound} \;
\nl \lIf {$\AlgVar{bound} \leq |\AlgVar{incumbent}|$}{\KwSty{return}} \label{PruneSearch}
\medskip
\nl $\langle \setG,\setH \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
\nl $v \gets \FuncSty{SelectVertex}(\setG)$ \label{SelectVertex} \;
\nl \For {$w \in \setH$ \label{WLoop}} {
    \nl    $M \gets M \cup \{(v,w)\}$ \label{GrowM} \;
\nl    \bf{Let} $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
\nl    \For {$\langle \setG',\setH'\rangle \in future$ \label{InnerLoop}}{
\nl        \bf{Let} $\setG'' \gets \setG' \cap \N(\graphG, v) \setminus \{v\}$ \label{NewPWithEdge} \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \N(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle \setG'' , \setH'' \rangle$ \label{AddToFutureWithEdge}}
\nl        \bf{Let} $\setG'' \gets \setG' \cap \invN(\graphG, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \invN(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} + \langle \setG'' , \setH'' \rangle$} \label{InnerLoopEnd}
       }
\nl   $\FuncSty{search}(\AlgVar{future'},M)$ \label{ExpandWithV} \;
\nl   $M \gets M \setminus \{(v,w)\}$ \label{ShrinkM} \;
  }
\nl $\setG' \gets \setG \setminus \{v\}$ \label{RemoveV} \;
\nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle \setG,\setH \rangle\}$\;
\nl \lIf {$\setG' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle \setG',\setH \rangle \}$}
\nl $\FuncSty{search}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
}
\;
\nl $\FuncSty{McSplit}(\graphG,\graphH)$ \label{McSplitFun} \;
\nl \Begin{
    \nl $\KwSty{global}~\AlgVar{incumbent} \gets \emptyset$ \;
\nl $\FuncSty{search}(\{\langle V(\graphG),V(\graphH) \rangle \},\emptyset)$ \label{FirstExpandCall} \;
\nl $\KwSty{return}$~$\AlgVar{incumbent}$ \;
}
\caption{Finding a maximum common subgraph.}
\label{McSplitAlg}
\end{algorithm}

@@

Our results so far suggest that \McSplit\ has broadly similar performance
trends to the constraint programming, forward-checking (CP-FC) algorithm of
\citet{DBLP:conf/cp/NdiayeS11}, but with much lower constant factors. Indeed,
in \cref{figure:prettyheatmaps1,figure:prettyheatmaps2} we plot the number of
recursive calls made by our algorithm versus the number made by
CP-FC, for the unlabelled and labelled, unconnected problem
instances. (Rather than a simple scatter plot, we use darker colours to
indicate a higher density of points around a location.) We see a close
correlation: we typically do slightly less work, and sometimes do more, but
instances with more than one order of magnitude difference in search space size
are rare.

@@

In the CP-FC algorithm, each vertex $v \in \V(\graphG)$ is represented by a
variable, whose domain corresponds to the set of vertices in $\V(\graphH)$ to which
$v$ may currently be mapped, with an additional special $\bot$ value
representing an unmapped vertex.
Given a label class $\langle \setG,\setH \rangle$ in \McSplit\, the vertices in $\setG$
correspond to variables in CP-FC. The label-class representation of domains is
possible because throughout the CP-FC algorithm for maximum common subgraph, the
domains of any two variables are either identical or disjoint (excluding
$\bot$, which is slightly more complicated). To the best of our knowledge, this
observation has not been made previously, and it is not exploited in other
implementations.

>>>>

We initially assume that graphs are unlabelled, undirected and without loops
(\cref{sec:extensions} describes how these restrictions may be relaxed).
The vertex and edge sets of a graph $\graphG$ are denoted $\V(\graphG)$ and $\E(\graphG)$.  The
set of vertices adjacent to vertex $v$ in graph $\graphG$ is called the
\emph{neighbourhood} of $v$, denoted $\N(\graphG, v)$. We denote by $\invN(\graphG, v)$ the
\emph{inverse neighbourhood} of $v$, being the set of the vertices not adjacent
to $v$ (excluding $v$ itself). A \emph{subgraph} of a graph $\graphG$ is a graph
consisting of some of the vertices of $\graphG$, and all of the edges between these
vertices. A \emph{common subgraph} of two graphs is a graph which is
(isomorphic to) a subgraph of two graphs simultaneously, and a \emph{maximum}
common subgraph is one with as many vertices as possible. (In this paper, all
subgraphs are induced. The \emph{maximum common partial subgraph} problem
instead asks for a common non-induced subgraph with as many \emph{edges} as
possible \citep{DBLP:conf/cp/NdiayeS11}.)

@@

\begin{algorithm}[t]
\DontPrintSemicolon
\nl $\FuncSty{search}(\AlgVar{future},M)$ \;
\nl \Begin{
%\nl \lIf {$\AlgVar{future} = \emptyset$ \bf{and} $|M| > |\AlgVar{incumbent}|$}
\nl \lIf {$|M| > |\AlgVar{incumbent}|$}{$\AlgVar{incumbent} \gets M$} \label{StoreIncumbent}
%\nl \lIf {$\AlgVar{future} = \emptyset$}{return}
\medskip
\nl $\AlgVar{bound} \gets |M|  + \sum_{\langle \setG,\setH \rangle \in \AlgVar{future}} \min(|\setG|,|\setH|)$ \label{CalcBound} \;
\nl \lIf {$\AlgVar{bound} \leq |\AlgVar{incumbent}|$}{\KwSty{return}} \label{PruneSearch}
\medskip
\nl $\langle \setG,\setH \rangle \gets \FuncSty{SelectLabelClass}(\AlgVar{future})$ \label{SelectClass} \;
\nl $v \gets \FuncSty{SelectVertex}(\setG)$ \label{SelectVertex} \;
\nl \For {$w \in \setH$ \label{WLoop}} {
    \nl    $M \gets M \cup \{(v,w)\}$ \label{GrowM} \;
\nl    \bf{Let} $\AlgVar{future'} \gets \emptyset$ \label{NewFuture} \;
\nl    \For {$\langle \setG',\setH'\rangle \in future$ \label{InnerLoop}}{
\nl        \bf{Let} $\setG'' \gets \setG' \cap \N(\graphG, v) \setminus \{v\}$ \label{NewPWithEdge} \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \N(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$ \label{AddToFutureWithEdge}}
\nl        \bf{Let} $\setG'' \gets \setG' \cap \invN(\graphG, v) \setminus \{v\}$ \label{NewPWithoutEdge}  \;
\nl        \bf{Let} $\setH'' \gets \setH' \cap \invN(\graphH, w) \setminus \{w\}$ \;
\nl        \If {$\setG'' \neq \emptyset$ \bf{and} $\setH'' \neq \emptyset$}{
\nl            $\AlgVar{future'} \gets \AlgVar{future'} \cup \{\langle \setG'' , \setH'' \rangle\}$} \label{InnerLoopEnd}
       }
\nl   $\FuncSty{search}(\AlgVar{future'},M)$ \label{ExpandWithV} \;
\nl   $M \gets M \setminus \{(v,w)\}$ \label{ShrinkM} \;
  }
\nl $\setG' \gets \setG \setminus \{v\}$ \label{RemoveV} \;
\nl $\AlgVar{future} \gets \AlgVar{future} \setminus \{\langle \setG,\setH \rangle\}$\;
\nl \lIf {$\setG' \neq \emptyset$} {$\AlgVar{future} \gets \AlgVar{future} \cup \{\langle \setG',\setH \rangle \}$}
\nl $\FuncSty{search}(\AlgVar{future},M)$ \label{ExpandWithoutV} \;
}
\;
\nl $\FuncSty{McSplit}(\graphG,\graphH)$ \label{McSplitFun} \;
\nl \Begin{
    \nl $\KwSty{global}~\AlgVar{incumbent} \gets \emptyset$ \;
\nl $\FuncSty{search}(\{\langle V(\graphG),V(\graphH) \rangle \},\emptyset)$ \label{FirstExpandCall} \;
\nl $\KwSty{return}$~$\AlgVar{incumbent}$ \;
}
\caption{Finding a maximum common subgraph.}
\label{McSplitAlg}
\end{algorithm}

@@

Our results so far suggest that \McSplit\ has broadly similar performance
trends to the constraint programming, forward-checking (CP-FC) algorithm of
\citet{DBLP:conf/cp/NdiayeS11}, but with much lower constant factors and memory usage. Indeed,
in \cref{figure:prettyheatmaps1,figure:prettyheatmaps2} we plot the number of
recursive calls made by our algorithm versus the number made by
CP-FC, for the unlabelled and labelled, unconnected problem
instances. (Rather than a simple scatter plot, we use darker colours to
indicate a higher density of points around a location.) We see a close
correlation: we typically do slightly less work, and sometimes do more, but
instances with more than one order of magnitude difference in search space size
are rare.

@@

In the CP-FC algorithm, each vertex $v \in \V(\graphG)$ is represented by a
variable, whose domain corresponds to the set of vertices in $\V(\graphH)$ to
which $v$ may currently be mapped, with an additional special $\bot$ value
representing an unmapped vertex.  Given a label class $\langle \setG,\setH
\rangle$ in \McSplit\, the vertices in $\setG$ correspond to variables in
CP-FC, and the vertices in $\setH$ to domain values. The label-class
representation of domains is possible because throughout the CP-FC algorithm
for maximum common subgraph, the domains of any two variables are either
identical or disjoint (excluding $\bot$, which is slightly more complicated).
To the best of our knowledge, this observation has not been made previously,
and it is not exploited in other implementations.
